---
title: "Data Science for Public Policy"
subtitle: "Assignment 04"
author: "Mitchell - mcz19 & Hannah Shepard-Moore - hms98"
execute:
  warning: false
format:
  html:
    embed-resources: true
---

The data used in this assignment comes from the US Department of Agriculture (USDA) Economic Research Service. The data and documentation can be found here: https://www.ers.usda.gov/data-products/supplemental-nutrition-assistance-program-snap-data-system/documentation/

Below is the code used to upload and read in this data. We specified the "County" tab because the Excel data came with several tabs including notes and state-level data.

```{r}
library(tidyverse)
library(readxl)
county_data_2_ <- read_excel("data/county_data (2).xls", 
    sheet = "County")
View(county_data_2_)
```

Our geographic locations of interest were Arizona and Virginia. In terms of variables of interest, these were the number of people using SNAP, the benefits used by those people, and the county population. We want to examine SNAP data across years as well as within years.

The code below was used to create the geographic subset:

```{r}
az_va <- county_data_2_ %>%  
  filter(STATE_CD== "AZ" | STATE_CD== "VA") %>%  
  mutate(POP00 = case_when(POP00 == "X" ~ NA))%>%
  select(CONAME, STATE_CD, METRO, URBCODE, 
         starts_with("POP0"), starts_with("PRGNUM0"), starts_with("PRGBEN0"))
```

Unfortunately, our data came in a "wide" format across years. So, in order to explore longitudinal data, we pivoted the variables of interest and merged them into a new dataset. 

The dataset did not have unique identifier for year and county, which we wanted to be our unit of analysis, so we created a unique variable for this within the "long" dataset for each variable of interest.

```{r}
#Population per year per county
pop_longer <- az_va %>% 
  select(CONAME, starts_with("POP0")) %>%
  pivot_longer(
    cols = starts_with("POP0"),
    names_to = "pop_year",
    values_to = "population",
    values_drop_na = TRUE) %>%
  mutate(year= case_when(pop_year == "POP01" ~ 2001,
                         pop_year == "POP02" ~ 2002,
                         pop_year == "POP03" ~ 2003,
                         pop_year == "POP04" ~ 2004,
                         pop_year == "POP05" ~ 2005,
                         pop_year == "POP06" ~ 2006,
                         pop_year == "POP07" ~ 2007,
                         pop_year == "POP08" ~ 2008,
                         pop_year == "POP09" ~ 2009,
                         pop_year == "POP10" ~ NA)) %>%
  mutate(CONAME = str_replace(CONAME,
                              " County, ",
                              ""),
         county_year = str_c(year,
                             CONAME,
                             sep= "_")) %>%
  select(county_year, population)

#Population on SNAP per year per county
snap_longer <- az_va %>%
  select(CONAME, starts_with("PRGNUM")) %>%
  pivot_longer(
    cols = starts_with("PRGNUM"),
    names_to = "snap_pop_year",
    values_to = "snap_pop",
    values_drop_na = TRUE) %>%
  mutate(year= case_when(snap_pop_year == "PRGNUM00" ~ 2000,
                         snap_pop_year == "PRGNUM01" ~ 2001,
                         snap_pop_year == "PRGNUM02" ~ 2002,
                         snap_pop_year == "PRGNUM03" ~ 2003,
                         snap_pop_year == "PRGNUM04" ~ 2004,
                         snap_pop_year == "PRGNUM05" ~ 2005,
                         snap_pop_year == "PRGNUM06" ~ 2006,
                         snap_pop_year == "PRGNUM07" ~ 2007,
                         snap_pop_year == "PRGNUM08" ~ 2008,
                         snap_pop_year == "PRGNUM09" ~ 2009)) %>%
  mutate(CONAME = str_replace(CONAME,
                              " County, ",
                              ""),
         county_year = str_c(year,
                             CONAME,
                             sep= "_")) %>%
  select(county_year, snap_pop)
#merge
snap_pop_longer <- merge(x= pop_longer,
                    y= snap_longer,
                    by= "county_year")
#SNAP Benefits per year per county
ben_longer <- az_va %>%
  select(CONAME, starts_with("PRGBEN")) %>%
  pivot_longer(
    cols = starts_with("PRGBEN"),
    names_to = "ben_year",
    values_to = "benefits",
    values_drop_na = TRUE) %>%
  mutate(year= case_when(ben_year == "PRGBEN00" ~ 2000,
                         ben_year == "PRGBEN01" ~ 2001,
                         ben_year == "PRGBEN02" ~ 2002,
                         ben_year == "PRGBEN03" ~ 2003,
                         ben_year == "PRGBEN04" ~ 2004,
                         ben_year == "PRGBEN05" ~ 2005,
                         ben_year == "PRGBEN06" ~ 2006,
                         ben_year == "PRGBEN07" ~ 2007,
                         ben_year == "PRGBEN08" ~ 2008,
                         ben_year == "PRGBEN09" ~ 2009))%>%
  mutate(CONAME = str_replace(CONAME,
                              " County, ",
                              ""),
         county_year = str_c(year,
                             CONAME,
                             sep= "_")) %>%
  select(county_year, benefits)
```

Upon creating a "long" dataset for each variable of interest, we merged these into one longitudinal dataset. Since we had to make a unique identifier for county and year to merge the long datasets, we brought back the year, state, and county variables here.

```{r}
#merge all + add back data
all_longer <- inner_join(x= snap_pop_longer,
                         y= ben_longer,
                         by= "county_year")
#bring back year by taking out first 4 digits of county_year var
all_longer <- all_longer %>% 
  mutate(year= substr(county_year, 1, 4)) %>%
  mutate(state= str_sub(county_year, -2))%>%
  mutate(county= str_sub(county_year, 6, -3))
```

We were interested in how the population on SNAP changed over the years.

```{r}
#snap pop by county
all_longer %>%
  filter(state == "AZ") %>%
  ggplot(mapping= aes(x= county,
                       y= snap_pop)) +
           geom_bar(color = "blue3", fill="blue3" ,stat = "identity", width = .5) +
  scale_y_continuous(name = "SNAP Recipients", breaks = c(100000, 200000, 300000, 400000, 500000, 600000, 700000, 800000, 900000, 1000000, 2000000, 3000000)) +
  labs(title = "SNAP Population by County in Arizona",
       subtitle = "Total SNAP Recipients by County from 2001-2009",
       x = "County" ,
       caption = "Data Source: USDA")
  theme_minimal()
```

Heatmap
```{r}
library(hrbrthemes)

all_longer %>% 
  filter(state == "AZ") %>%
  mutate(snap_ratio = snap_pop/population) %>%
  ggplot(aes(x = year, y = county, fill = snap_ratio)) +
  geom_tile() +
  scale_fill_gradient(low = "white", high = "blue4") +
  labs(title = "Proportion of County Residents on SNAP in Arizona",
       subtitle = "A Greater Proportion of Apache and Navajo County Residents are on SNAP ",
       x = "Year",
       y = "County",
       caption = "Date Source: USDA")
  theme_light()
```


Since we live in Northern VA, we explored the history of SNAP benefits in this area. One question we had was: "How did SNAP benefits change around the time of the 2008 recession?". The below graph shows how benefits trended before the recession in 5 'NOVA' counties and how they changed after 2008.
```{r}
#snap use over time
all_longer %>%
  filter(state == "VA",
         snap_pop >= 0,
         population > 0,
         county== "Arlington" | 
           county== "Fairfax" |
           county== "Fauquier" |
           county== "Loudoun" |
           county== "Prince William") %>%
  mutate(ben_share= benefits / population) %>%
  ggplot(mapping= aes(x= year,
                      y= ben_share, 
                      color= county))+
  geom_point()+
  geom_line(aes(group= county))+
  theme_light()+
  scale_color_manual(name= "",
                     values= c("Arlington" = "brown1",
                               "Fairfax" = "goldenrod1",
                               "Fauquier"= "darkolivegreen1",
                               "Loudoun"= "dodgerblue2",
                               "Prince William"= "mediumorchid2"),
                     labels= c("Arlington" = "Arlington County",
                               "Fairfax" = "Fairfax County",
                               "Loudoun" = "Loudon County",
                               "Fauquier" = "Fauquier County",
                               "Prince William" = "Prince William County"))+
  labs(title= "Northern VA counties saw a spike in SNAP benefits per capita after 2008",
       subtitle= "SNAP benefits were trending slightly up before the recession started.",
       caption= "Source: US Department of Agriculture",
       x= "",
       y= "SNAP benefits per capita")

```

Box Graph
```{r}
all_longer %>%
  ggplot(aes(x = state, y = benefits, fill = state)) + 
  geom_boxplot() +
  geom_jitter() +
  labs(title = "Arizona Paid More in Benefits than Virginia Between 2001-2009",
       subtitle = "Arizona Counties had Substantially More Outliers Compared to the State Average",
       x ="State" ,
       y = "SNAP Benefits",
       caption = "Date Source: USDA")
```

